---
title: "Building 1023-EZ Open Database"
output:
  html_document:
    theme: united
    df_print: paged
    highlight: tango
    smart: false
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, fig.width=8)
```


In this step we will geocode addresses of the founding teams (managers and board members) and the nonprofit addresses.


# Packages

```{r}
library( dplyr )
library( tidyr )
library( httr )
```



```{r}
dat.people <- readRDS( "Data/PEOPLE-2014-2018.rds" )
head( dat.people, 20 )

dat.npos <- readRDS( "Data/NONPROFIT-ADDRESSES-2014-2018.rds" )
head( dat.npos,20 )
```




# Geocode Demo


**CENSUS GEOCODER** 

DOCUMENTATION:
https://www.census.gov/geo/maps-data/data/geocoder.html

WEB FORM:
https://geocoding.geo.census.gov/geocoder/geographies/addressbatch?form

Addresses should be formatted in a single line with comma delimiters. The address should consist of:

* Unique ID,
* House Number and Street Name,
* City,
* State,
* ZIP Code


```{r}
addr <- select( dat.people, ID, Ofcrdirtruststreetaddr, Ofcrdirtrustcity, Ofcrdirtruststate, Ofcrdirtrustzip )
names( addr ) <- c("ID", "STREET", "CITY", "STATE", "ZIP" )

addr.test <- addr[ 1:20, ]
write.csv( addr.test, "TestAddresses.csv", row.names=F )



### RETURN GEOGRAPHIES (lat/lon plus census tracts & blocks)
### library( httr )

apiurl <- "https://geocoding.geo.census.gov/geocoder/geographies/addressbatch"

addressFile <- "TestAddresses.csv"

resp <- POST( apiurl, 
              body=list(addressFile=upload_file(addressFile), 
                        benchmark="Public_AR_Census2010",
                        vintage="Census2010_Census2010",
                        returntype="csv" ), 
              encode="multipart" )

# content( resp, as="text" )
# content( resp )

results <- "results.csv"
var_names <- c( "id", "input_address", 
                "match", "match_type", 
                "out_address", "lat_lon", 
                "tiger_line_id", "tiger_line_side", 
                "state_fips", "county_fips", 
                "tract_fips", "block_fips" )
v.names <- paste(var_names, collapse=',')
writeLines( text=c(v.names, content(resp)) , con="results.csv" )



### ADD VARIABLE NAMES
# from: https://www.census.gov/geo/maps-data/data/geocoder.html

res <- read.csv( "results.csv", header=T, stringsAsFactors=F, colClasses="character" )


### SPLIT LATITUDE AND LONGITUDE COORDINATES

lat.lon <- strsplit( res$lat_lon, "," )

for( i in 1:length(lat.lon) )
{
   # print( length( lat.lon[[i]] ) )
   if( length( lat.lon[[i]] ) < 2 )
   lat.lon[[ i ]] <- c(NA,NA) 
}

m <- matrix( unlist( lat.lon ), ncol=2, byrow=T )
colnames(m) <- c("lon","lat")
m <- as.data.frame( m )

res <- cbind( res, m )
head( res )

# write.csv( res, "ResultsAugmented.csv", row.names=F )
```









# Batch Geocode People


```{r, eval=F}
dat.people <- readRDS( "Data/PEOPLE-2014-2018.rds" )
```


"Batch files may not exceed 10,000 records."



```{r}

d2 <- dat.people

# Geocode input file variables: 
# ID	STREET	CITY	STATE	ZIP

dat.people <- select( dat.people, ID, Ofcrdirtruststreetaddr, 
              Ofcrdirtrustcity, Ofcrdirtruststate, Ofcrdirtrustzip )

dir.create( "addresses_people" )
setwd( "addresses_people" )

names( dat.people ) <- NULL  # input file should not have names

# split address files into files with 500 addresses each
loops <- ceiling( nrow( dat.people ) / 500 )


# SPLIT DATA INTO PARTS
for( i in 1:loops )
{

filename <- paste0( "AddressesPeople",i,".csv" )

start.row <- ((i-1)*500+1)
end.row <- (500*i)
if( nrow(dat.people) < end.row ){ end.row <- nrow(dat.people) }

write.csv( dat.people[ start.row:end.row, ], filename, row.names=F )
print( i )
print( paste( "Start Row:", start.row ) )
print( paste( "End Row:", end.row ) )

}



### START GEOCODING
### RETURN GEOGRAPHIES (lat/lon plus census tracts & blocks)

for( i in 1:loops )
{

filename <- paste0( "AddressesPeople",i,".csv" )

addressFile <- filename

apiurl <- "https://geocoding.geo.census.gov/geocoder/geographies/addressbatch"


print( i )
start_time <- Sys.time()          ###############

try( 
resp <- POST( apiurl, 
              body=list(addressFile=upload_file(addressFile), 
                        benchmark="Public_AR_Census2010",
                        vintage="Census2010_Census2010",
                        returntype="csv" ), 
              encode="multipart" )

# content( resp )
)

end_time <- Sys.time()
print( end_time - start_time )    

################



### ADD VARIABLE NAMES
# from: https://www.census.gov/geo/maps-data/data/geocoder.html


filename2 <- paste0( "results",i,".csv" )

results <- filename2

var_names <- c( "id", "input_address", 
                "match", "match_type", 
                "out_address", "lat_lon", 
                "tiger_line_id", "tiger_line_side", 
                "state_fips", "county_fips", 
                "tract_fips", "block_fips" )
v.names <- paste(var_names, collapse=',')
writeLines( text=c(v.names, content(resp)) , con=filename2 )

# writeLines( text=content( resp ) , con=filename2 )  




res <- read.csv( filename2, header=T, 
                 stringsAsFactors=F, 
                 colClasses="character" )


### SPLIT LATITUDE AND LONGITUDE COORDINATES

lat.lon <- strsplit( res$lat_lon, "," )

for( j in 1:length(lat.lon) )
{
   # print( length( lat.lon[[j]] ) )
   if( length( lat.lon[[j]] ) < 2 )
   lat.lon[[ j ]] <- c(NA,NA) 
}

m <- matrix( unlist( lat.lon ), ncol=2, byrow=T )
colnames(m) <- c("lon","lat")
m <- as.data.frame( m )

res <- cbind( res, m )

write.csv( res, paste0("ResultsPeople",i,".csv"), row.names=F )



}  # end of loop


```



### Combine Files

```{r}
dir()

these <- (dir())[ grepl( "ResultsPeople", dir() ) ]

dat <- read.csv( these[1], stringsAsFactors=F )


for( i in 2:length(these) )
{

  d <- read.csv( these[i], stringsAsFactors=F )
  dat <- bind_rows( dat, d ) 

}

plot( dat$lon, dat$lat, pch=19, cex=0.5 )

plot( dat$lon, dat$lat, pch=19, cex=0.5, col=gray(0.5,0.01) )

getwd()

# write.csv( dat, "PEOPLE-GEOCODES.csv", row.names=F )

saveRDS( dat, "../Data/PEOPLE-GEOCODES.rds" )

setwd( ".." )  # return to main directory
```






# Batch Geocode Nonprofits

copy code for people, adapt for nonprofit addresses



```{r, eval=F}
dat.npos <- readRDS( "Data/NONPROFIT-ADDRESSES-2014-2018.rds" )
```






```{css, echo=F, eval=T}
p {
  color: black;
  font-size:1.2em;
  margin: 20px 0 20px 0 !important;
}

p.caption {
  text-align: center;
  font-weight: bold;
}

th { font-weight: bold; }

td {
    padding: 3px 10px 3px 10px !important;
    text-align: center;
}

table
{ 
    margin-left: auto;
    margin-right: auto;
    margin-top:80px;
    margin-bottom:100px;
}

h1, h2, h3{
  margin-top:100px !important;
  margin-bottom:20px !important;
}

h5{
    text-align: center;
    color: gray;
    font-size:0.8em;
}

img {
    max-width: 90%;
    display: block;
    margin-right: auto;
    margin-left: auto;
    margin-top:30px !important;
    margin-bottom:30px !important;
}


.sourceCode {
   margin-top:50px;
}

.pagedtable-wrapper {
   margin-bottom:30px;
}
```
